### 2023-03-02 calculus.tex
Updates from Overleaf

Programmers often work with program states that are syntactically well-formed but not well-typed.
\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
environment. For example, a simple error that may occur is the use of an unbound variable. As in
\cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
invalid by highlighting it in red.

@@

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the conditional are of different types: $5$ is a number, while
$\textsf{false}$ is a boolean. Hazel opts to highlight the entire expression in such a situation,
while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
the other.

@@

We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with explicit analysis rules. In other words, we restrict subsumption to
be the rule of ``last resort'', which is necessary to establish that marking and typing are
deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

@@

To satisfy marking totality, such a possibility motivates an \emph{inconsistent type mark}
$\ECInconType{\ECMV}$, which states exactly that the synthesized type of $\ECMV$ is inconsistent
with the expected type. It is then accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{MKAInconsistentTypes}

@@

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  %
  \caption{Matched arrow types.}
  \label{fig:calculus-matched-arrow}
\end{figure}

@@

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECInconAsc{x}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV_1}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam3}
\end{mathpar}

@@

In the unmarked language, only a synthesis rule is necessary for applications:
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynTypeU{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology, we can straightforwardly give the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{MKSAp1}

@@

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as products. Indeed, the same approach is taken for projections in
\cref{sec:calculus-products}. Another approach might indicate the same kind of error but mark the
entire application.

@@

The boolean values are straightforward to handle, behaving in a similar manner as numbers: 
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ETrue}{\TBool}
  }{USTrue}

@@

Conditionals, however, offer an interesting case. In the unmarked language, we have both explicit
synthetic and analytic rules:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\
    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{USIf}

@@

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
Similar to previous cases, we recursively mark the guard and branches, constructing the marked
conditional from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{MKSIf}

@@

The framework also allows for other localization approaches. For example, as discussed at the
beginning of the section, we may regard the first branch as ``correct'' and localize any errors to
the second. The following rule formalizes such a design:
\[%
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKSIf'}
\]%
Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
heuristics may be devised for different kinds of branching structures.

\subsubsection{Holes}
For completeness, we finish the development of the marked lambda calculus on the extended GTLC with
the marking of holes. Like number literals and booleans, these are governed by straightforward
rules.
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\EEHole}{\TUnknown}
  }{USHole}

@@

\subsubsection{Additional Metatheorems}
To conclude this section, we present two more metatheorems that help ensure correctness of the
system. Though \cref{thm:calculus-marking-well-formedness} guarantees that marking does not change
the syntactic structure of a program, it makes no statement about the presence of error marks in the
resulting marked program. Following the logic of constructing marking rules, we have
\cref{thm:calculus-marking-well-ill-typed}.

@@

    \item \begin{enumerate}
        \item If there does not exist $\TMV$ such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
        \item If there does not exist $\TMV$ such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
      \end{enumerate}
  \end{enumerate}
\end{theorem}
%
That is, since error marks indicate type errors in the program, no marks should be introduced
when marking a well-typed program, a property given by the judgment $\markless{\ECMV}$. The
converse---that at least one mark is inserted into ill-typed programs---is also true.

Finally, no less importantly, marking is deterministic. This is given by
\cref{thm:calculus-marking-unicity}.
%
\begin{theorem}[name=Marking Unicity] \
  \label{thm:calculus-marking-unicity}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
      \TMV_2$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV}$ and
      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV}$, then $\ECMV_1 = \ECMV_2$.
  \end{enumerate}
\end{theorem}
%
Together, totality and unicity give that marking may be considered and implemented as a total
function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to extract
an implementation from these rules.

@@

We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
by adding binary product types, written $\TProd{\TMV}{\TMV}$. In the interest of space, we focus on
aspects related to marking; a comprehensive set of rules is found in the supplemental material.

First, we extend the syntax of types and the accompanying notions of consistency and join, which
closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
arrow types, the judgment $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be
considered the binary product type $\TProd{\TMV_1}{\TMV_2}$ [cite?].\todo{cite}

The unmarked language is extended with a pair constructor $\EPair{\EMV}{\EMV}$ and projection
operators $\EProjL{\EMV}$ and $\EProjR{\EMV}$, governed by standard bidirectional rules. In addition
to the synthesis rules, an explicit analysis rules is defined for the former: if an expected type is
known, the responsibility for type errors is shift to the components.

The marked language is extended with the same forms, and we again devise marking rules by modelling
off the unmarked typing rules.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }{MKSPair}

@@

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
  }{MKSProjR1} \\

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair1}
\end{mathpar}

However, as previously seen with lambda abstractions and applications, two sets of rules are further
necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
binary product may fail, requiring the introduction of a new error mark,
$\ECAnaNonMatchedProd{\ECMV}$, which is analogous to $\ECAnaNonMatchedArrow{\ECMV}$ for lambda
abstractions. Another kind of error may arise when the subject of a projection does not synthesize a
matched product type. As in our development of marking for applications, we utilize a new mark,
written $\ECSynNonMatchedProd{\ECMV}$. These cases give the final marking rules for products:
%
\begin{mathpar}
  \judgment{
    \notMatchedProd{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair2} \\

>>>>

Programmers often work with program states that are syntactically well-formed but not well-typed.
\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
environment. For example, a simple error that may occur is the use of an unbound variable (\cref{fig:calculus-examples-unbound}).
% As in
% \cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
% invalid by highlighting it in red.

@@

With local type inference, an expression might have a expected type. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates an inconsistency between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error caused by inconsistent
branch types. As discussed in the introduction, Hazel opts to highlight the entire expression in such a situation,
while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
the other.

@@

We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with both synthesis and analysis rules (see below). In other words, we restrict subsumption to
be the rule of ``last resort'', which is necessary to establish that marking and typing are
deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

@@

To satisfy marking totality, such a possibility motivates an \emph{inconsistent type mark}
$\ECInconType{\ECMV}$, which is applied when the synthesized type of $\ECMV$ is inconsistent
with the expected type:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{MKAInconsistentTypes}

@@

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  \vspace{-10px}
  \caption{Matched arrow types.}
  \label{fig:calculus-matched-arrow}
\end{figure}

@@

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECInconAsc{x}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV_1}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam3}
\end{mathpar}

@@

In the unmarked language, only a synthesis rule is necessary for applications:
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynTypeU{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology, we have the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{MKSAp1}

@@

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as products. Indeed, the same approach is taken for projections in
\cref{sec:calculus-products}. 
% Another approach might indicate the same kind of error but mark the
% entire application.

@@

The boolean values are similar to numbers: 
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ETrue}{\TBool}
  }{USTrue}

@@

Conditionals, however, are an interesting case. In the unmarked language, we have both explicit
synthetic and analytic rules:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\
    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{USIf}

@@

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
% Similar to previous cases, we recursively mark the guard and branches, constructing the marked
% conditional from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{MKSIf}

@@

The framework also allows for other localization approaches. For example, as discussed at the
beginning of the section, we may regard the first branch as ``correct'' and localize any errors to
the second. The following rule formalizes such a design:
\[%
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKSIf'}
\]%
% Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
% heuristics may be devised for different kinds of branching structures.

\subsubsection{Holes}
For completeness, we finish the development of the marked lambda calculus on the extended GTLC with
the marking of empty holes, which are never marked.
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\EEHole}{\TUnknown}
  }{USHole}

@@

\subsubsection{Additional Metatheorems}
To conclude this section, we present two more metatheorems that help ensure correctness of the
system. Though \cref{thm:calculus-marking-well-formedness} guarantees that marking does not change
the syntactic structure of a program, it makes no statement about the presence of error marks in the
resulting marked program. 
\cref{thm:calculus-marking-well-ill-typed} established that well-typed expressions are left unmarked and ill-typed expressions have at least one mark.

@@

    \item \begin{enumerate}
        \item If there does not exist $\TMV$ such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
        \item If there does not exist $\TMV$ such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
      \end{enumerate}
  \end{enumerate}
\end{theorem}
%
Finally, no less importantly, marking is deterministic. This is given by
\cref{thm:calculus-marking-unicity}.
%
\begin{theorem}[name=Marking Unicity] \
  \label{thm:calculus-marking-unicity}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
      \TMV_2$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV}$ and
      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV}$, then $\ECMV_1 = \ECMV_2$.
  \end{enumerate}
\end{theorem}
%
Together, totality and unicity give that marking may be implemented as a total
function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to implement these rules.

@@

The marked lambda calculus is intended as a minimal core. The methodology described above can be used to extend the marked lambda calculus with new typing features. For example, let us add binary product types, written $\TProd{\TMV_1}{\TMV_2}$. In the interest of space, we focus on
aspects related to marking and omit standard rules; a comprehensive set of rules is found in the supplemental material.

% First, we extend the syntax of types and the accompanying notions of consistency and join, which
% closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
% arrow types, we need a matched product type relation $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$.

The unmarked language is extended with a pair constructor $\EPair{\EMV}{\EMV}$ and projection
operators $\EProjL{\EMV}$ and $\EProjR{\EMV}$, governed by standard bidirectional rules. In addition
to the synthesis rules, an explicit analysis rules is defined for pairs: if the expected type has a matched product type (cf. matched arrow types above), the responsibility for type errors is shifted to the components.

We can devise the basic marking rules from the unmarked typing rules.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }{MKSPair}

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair1}

@@

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
  }{MKSProjR1}
\end{mathpar}
However, as previously seen with lambda abstractions and applications, two sets of rules are further
necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
binary product may fail, requiring the introduction of a new error mark,
$\ECAnaNonMatchedProd{\ECMV}$, which is analogous to $\ECAnaNonMatchedArrow{\ECMV}$ for lambda
abstractions. Another kind of error may arise when the subject of a projection does not synthesize a
matched product type. As in our development of marking for applications, we utilize a new mark,
written $\ECSynNonMatchedProd{\ECMV}$. These cases give the final marking rules for products:
%
\begin{mathpar}
  \judgment{
    \notMatchedProd{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair2} \\

### 2023-03-02 let.tex
Updates from Overleaf

\subsection{Destructuring Let with Composite Patterns}
\label{sec:calculus-let}

To ease the use of products, many languages feature destructuring bindings [cite]. Let us add let
expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. In a bidirectional setting, the pattern
itself synthesizes a type which constrains the definition.
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%
It turns out, however, that these composite patterns with type annotations can complicate type error
localization in subtle ways.

@@

Though the distinction is benign in this example, there is a subtle semantic distinction: synthetic
mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type
constraint. This manifests when expressions may have internal type inconsistencies, such as in the
following program:
\[%
  \ELet{\PVar{a}}{\EIf{\ETrue}{1}{\EFalse}}{\EMV}
\]%
If the pattern $\PVar{a}$ suggests that the conditional is in synthetic position, it will be marked
with an inconsistent branches error following our development of marking above. If instead it is in
analytic position against the unknown type, no mark will be produced.

@@

This system is achieved by augmenting the type system with a variant of the unknown type, written
$\TUnknownSwitch$, that triggers a ``switch'' to synthesis and otherwise behaves identically to
$\TUnknown$. In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
$\TUnknownSwitch$, and the annotated pattern $\PAsc{\PVar{b}}{\TUnknown}$ synthesizes $\TUnknown$.
We write $\ctxSynPatU{\ctx}{\PMV}{\TMV}$ to say that the pattern $\PMV$ synthesizes type $\TMV$.
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\ctx}{\EMV}{\ctx'}
  }{
    \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
  }{UASynSwitch}

@@

To ensure propagation of type information between pattern, definition, and body, the destructuring
let requires some extra machinery.
\[%
  \judgment{
    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
  }{USLetPat}
\]%
First, to ensure that the pattern and definition types are consistent, the pattern synthesizes a
type, against which the definition is analyzed. Then, the definition synthesizes a type, against
which the pattern is analyzed. This analysis is guaranteed to succeed; we need the context it
produces for use in the synthesis of the body's type. The analytic rule is similar, with the final
premise and conclusion changed to analysis.

>>>>

\subsection{Destructuring Let with Type Annotated Composite Patterns}
\label{sec:calculus-let}

To ease the use of products, many languages feature destructuring bindings. In a typed setting, we may want to granularly add type annotations in patterns as well. As it turns out this combination of features is surprisingly tricky to get right! Let us add let
expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

Though the distinction is benign in this example, there is a subtle semantic distinction: synthetic
mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type
constraint. This manifests when expressions may have internal type inconsistencies,  as in the
following:
\[%
  \ELet{\PVar{a}}{\EIf{\ETrue}{1}{\EFalse}}{\EMV}
\]%
If the pattern $\PVar{a}$ suggests that the conditional is in synthetic position, it will be marked
with an inconsistent branches error following our development of marking above. If instead it is in
analytic position against the unknown type, no mark will be produced.

@@

This system is achieved by augmenting the type system with a variant of the unknown type, written
$\TUnknownSwitch$, that triggers a ``switch'' to synthesis and otherwise behaves identically to
$\TUnknown$. In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
$\TUnknownSwitch$, and the annotated pattern $\PAsc{\PVar{b}}{\TUnknown}$ synthesizes $\TUnknown$.
We write $\ctxSynPatU{\ctx}{\PMV}{\TMV}$ to say that the pattern $\PMV$ synthesizes type $\TMV$.
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV}{\ctx'}
  }{
    \ctxAnaTypeU{\ctx}{\EMV}{\TUnknownSwitch}
  }{UASynSwitch}

@@

To ensure propagation of type information between pattern, definition, and body, the destructuring
let requires some extra machinery.
\[%
  \judgment{
    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
  }{USLetPat}
\]%
First, to ensure that the pattern and definition types are consistent, the pattern synthesizes a
type, against which the definition is analyzed. Then, the definition synthesizes a type, against
which the pattern is analyzed. This analysis is guaranteed to succeed; we need the context it
produces for use in the synthesis of the body's type. The analytic rule is similar, with the final
premise and conclusion changed to analysis.
